---
title: "Assignment 1 - Visualisation"
author: "Namesh Panneer Selvan (22148363)"
date: "2026-07-24"
output:
  word_document: default
  pdf_document: default
---

# DECLARATION
I hold a copy of this assignment that I can produce if the original is lost or damaged. 

I hereby certify that no part of this assignment/product has been copied from any other student’s work or from any other source except where due acknowledgement is made in the assignment.

No part of this assignment/product has been written/produced for me by another person except
where such collaboration has been authorised by the subject lecturer/tutor concerned.

# 1. Required Packages
```{r}
required_packages <- c(
  "igraph", "visNetwork", "ggraph", "ggplot2",
  "dplyr", "readr", "htmlwidgets", "scales"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

library(igraph)
library(visNetwork)
library(ggraph)
library(ggplot2)
library(dplyr)
library(readr)
library(htmlwidgets)
library(scales)

set.seed(42)
```

# 2. Load and validate data
```{r}
nodes_file <- "crisis-nodes.csv"
edges_file <- "crisis-edges.csv"

if (!file.exists(nodes_file) || !file.exists(edges_file)) {
  stop(
    "The files 'crisis-nodes.csv' and 'crisis-edges.csv' must be in the working directory."
  )
}

nodes_raw <- read_csv(nodes_file, show_col_types = FALSE)
edges_raw <- read_csv(edges_file, show_col_types = FALSE)

required_node_columns <- c("Id", "Label")
required_edge_columns <- c("Source", "Target", "Weight")

if (!all(required_node_columns %in% names(nodes_raw))) {
  stop("The node file must contain the columns: Id and Label.")
}

if (!all(required_edge_columns %in% names(edges_raw))) {
  stop("The edge file must contain the columns: Source, Target and Weight.")
}

nodes <- nodes_raw %>%
  transmute(
    name = as.character(Id),
    label = as.character(Label),
    type = if ("d1" %in% names(nodes_raw)) as.character(d1) else "author"
  ) %>%
  filter(!is.na(name), name != "") %>%
  distinct(name, .keep_all = TRUE)

edges <- edges_raw %>%
  transmute(
    from = as.character(Source),
    to = as.character(Target),
    weight = suppressWarnings(as.numeric(Weight))
  ) %>%
  filter(
    !is.na(from), from != "",
    !is.na(to), to != "",
    from != to
  ) %>%
  mutate(weight = if_else(is.na(weight) | weight <= 0, 1, weight)) %>%
  group_by(from, to) %>%
  summarise(weight = sum(weight), .groups = "drop")

unknown_vertices <- setdiff(unique(c(edges$from, edges$to)), nodes$name)
if (length(unknown_vertices) > 0) {
  warning(
    "Some edge endpoints were not present in the node file and were added automatically: ",
    paste(unknown_vertices, collapse = ", ")
  )
  nodes <- bind_rows(
    nodes,
    tibble(name = unknown_vertices, label = unknown_vertices, type = "author")
  )
}

cat("Loaded", nrow(nodes), "nodes and", nrow(edges), "unique edges.\n")

```

# 3. Build the undirected weighted graph
```{r}
g <- graph_from_data_frame(
  d = edges,
  directed = FALSE,
  vertices = nodes
)

g <- simplify(
  g,
  remove.multiple = TRUE,
  remove.loops = TRUE,
  edge.attr.comb = list(weight = "sum", "ignore")
)

# Edge weight represents collaboration strength. For weighted shortest-path
# measures, stronger ties should correspond to shorter distances.
E(g)$distance <- 1 / E(g)$weight
```

# 4. Network metrics
```{r}
V(g)$degree <- degree(g, mode = "all")
V(g)$weighted_degree <- strength(g, mode = "all", weights = E(g)$weight)

# Unweighted measures are easier to interpret for this assignment.
V(g)$betweenness <- betweenness(g, directed = FALSE, normalized = TRUE)
V(g)$closeness <- closeness(g, normalized = TRUE)
V(g)$eigenvector <- eigen_centrality(g, directed = FALSE, weights = E(g)$weight)$vector

# Weighted alternatives, using inverse collaboration strength as distance.
V(g)$weighted_betweenness <- betweenness(
  g,
  directed = FALSE,
  weights = E(g)$distance,
  normalized = TRUE
)
V(g)$weighted_closeness <- closeness(
  g,
  weights = E(g)$distance,
  normalized = TRUE
)

# Louvain requires an undirected graph and positive edge weights.
community_model <- cluster_louvain(g, weights = E(g)$weight)
V(g)$community <- membership(community_model)

network_summary <- tibble(
  metric = c(
    "Nodes", "Edges", "Density", "Components", "Average path length",
    "Diameter", "Global clustering coefficient", "Communities", "Modularity"
  ),
  value = c(
    vcount(g),
    ecount(g),
    round(edge_density(g, loops = FALSE), 4),
    components(g)$no,
    round(mean_distance(g, directed = FALSE), 3),
    diameter(g, directed = FALSE),
    round(transitivity(g, type = "global"), 3),
    length(community_model),
    round(modularity(community_model), 3)
  )
)

print(network_summary)
write_csv(network_summary, "crisis_network_summary.csv")

metrics <- tibble(
  author = V(g)$label,
  degree = V(g)$degree,
  weighted_degree = V(g)$weighted_degree,
  betweenness = V(g)$betweenness,
  closeness = V(g)$closeness,
  eigenvector = V(g)$eigenvector,
  weighted_betweenness = V(g)$weighted_betweenness,
  weighted_closeness = V(g)$weighted_closeness,
  community = V(g)$community
) %>%
  arrange(desc(degree), desc(weighted_degree))

cat("\nTop 10 authors by degree centrality:\n")
print(metrics %>% select(author, degree, weighted_degree, betweenness, community) %>% head(10))

cat("\nTop 10 authors by betweenness centrality:\n")
print(
  metrics %>%
    arrange(desc(betweenness)) %>%
    select(author, degree, weighted_degree, betweenness, community) %>%
    head(10)
)

write_csv(metrics, "crisis_metrics.csv")
```

# 5. Interactive network visualisation
```{r}
community_palette <- c(
  "#4E79A7", "#59A14F", "#E15759", "#F1CE63",
  "#B07AA1", "#76B7B2", "#FF9DA7", "#9C755F"
)

vertex_table <- igraph::as_data_frame(
  g,
  what = "vertices"
)

edge_table <- igraph::as_data_frame(
  g,
  what = "edges"
)

vis_nodes <- vertex_table %>%
  transmute(
    id = name,
    label = label,
    value = degree,
    group = as.character(community),
    title = paste0(
      "<b>", label, "</b>",
      "<br>Degree: ", degree,
      "<br>Weighted degree: ", round(weighted_degree, 1),
      "<br>Betweenness: ", round(betweenness, 4),
      "<br>Eigenvector centrality: ", round(eigenvector, 4),
      "<br>Community: ", community
    )
  )

vis_edges <- edge_table %>%
  transmute(
    from = from,
    to = to,
    value = weight,
    title = paste0("Collaboration weight: ", weight)
  )

interactive_network <- visNetwork(
  vis_nodes,
  vis_edges,
  main = "The Crisis Co-authorship Network",
  width = "100%",
  height = "800px"
) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visGroups(groupname = "1", color = community_palette[1]) %>%
  visGroups(groupname = "2", color = community_palette[2]) %>%
  visGroups(groupname = "3", color = community_palette[3]) %>%
  visGroups(groupname = "4", color = community_palette[4]) %>%
  visGroups(groupname = "5", color = community_palette[5]) %>%
  visNodes(
    scaling = list(min = 8, max = 45),
    font = list(size = 14, face = "Arial")
  ) %>%
  visEdges(
    color = list(color = "#CCCCCC", highlight = "#333333"),
    scaling = list(min = 1, max = 8),
    smooth = FALSE
  ) %>%
  visOptions(
    highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE),
    nodesIdSelection = list(enabled = TRUE, useLabels = TRUE),
    selectedBy = list(variable = "group", multiple = TRUE)
  ) %>%
  visInteraction(
    navigationButtons = TRUE,
    keyboard = TRUE,
    tooltipDelay = 100,
    hover = TRUE
  ) %>%
  visPhysics(stabilization = list(enabled = TRUE, iterations = 1000)) %>%
  visLayout(randomSeed = 42)

visSave(
  interactive_network,
  file = "crisis_interactive.html",
  selfcontained = TRUE
)
```

# 6. Static network overview
```{r}
network_plot <- ggraph(g, layout = "fr") +
  geom_edge_link(
    aes(width = weight),
    colour = "grey80",
    alpha = 0.65,
    show.legend = FALSE
  ) +
  geom_node_point(
    aes(size = degree, colour = factor(community)),
    alpha = 0.95
  ) +
  geom_node_text(
    aes(label = ifelse(degree >= 15, label, "")),
    repel = TRUE,
    size = 3.2,
    colour = "black"
  ) +
  scale_edge_width(range = c(0.2, 2.5)) +
  scale_size_continuous(range = c(2, 14), name = "Degree") +
  scale_colour_manual(
    values = community_palette,
    name = "Community"
  ) +
  labs(
    title = "The Crisis Co-authorship Network",
    subtitle = "Node size represents degree; colour represents Louvain community"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  )

ggsave(
  filename = "fig_network_overview.png",
  plot = network_plot,
  width = 11,
  height = 8,
  dpi = 300,
  bg = "white"
)

```

# 7. Top 15 authors by degree
```{r}
# Top 15 authors by degree
top_15 <- metrics %>%
  arrange(desc(degree), desc(weighted_degree)) %>%
  slice_head(n = 15) %>%
  mutate(
    community = factor(community),
    author = factor(author, levels = rev(author))
  )

print(top_15 %>% select(author, degree, community))

top_degree_plot <- ggplot(
  top_15,
  aes(x = author, y = degree, fill = community)
) +
  geom_col(width = 0.8) +
  geom_text(
    aes(label = degree),
    hjust = -0.15,
    size = 3.5
  ) +
  coord_flip(clip = "off") +
  scale_fill_manual(
    values = community_palette[seq_len(nlevels(top_15$community))],
    name = "Community"
  ) +
  scale_y_continuous(
    limits = c(0, max(top_15$degree) * 1.12),
    breaks = scales::pretty_breaks()
  ) +
  labs(
    title = "Top 15 Authors by Degree Centrality",
    subtitle = "Higher degree indicates collaboration with more authors",
    x = NULL,
    y = "Degree"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    panel.grid.major.y = element_blank(),
    legend.position = "right",
    plot.margin = margin(5.5, 20, 5.5, 5.5)
  )

print(top_degree_plot)

ggsave(
  "fig_top_degree.png",
  plot = top_degree_plot,
  width = 10,
  height = 7,
  dpi = 300,
  bg = "white"
)
```

# 8. Adjacency matrix visualisation
```{r}
# A second, contrasting representation of the SAME network.

# Only dependency is `g` with its degree and community attributes (Section 4).
 
# --- Subset to the most connected authors ---------------------------------
# A full 96x96 matrix with readable labels will not fit on a report page.
# The top 25 by degree keeps every community represented and legible.
top_n_authors <- 25
keep_idx <- order(V(g)$degree, decreasing = TRUE)[seq_len(min(top_n_authors, vcount(g)))]
gs <- induced_subgraph(g, keep_idx)

# --- Order rows and columns by community, then degree ----------------------
# THIS IS THE ANALYTICAL STEP. An arbitrarily ordered matrix looks like noise;
# ordering by community pushes dense blocks onto the diagonal, so the structure
# the force-directed layout found reappears as visible squares.
ord        <- order(V(gs)$community, -V(gs)$degree)
lev        <- V(gs)$label[ord]
comm_order <- V(gs)$community[ord]
 
# --- Weighted adjacency matrix in long format ------------------------------
adj <- as_adjacency_matrix(gs, attr = "weight", sparse = FALSE)
dimnames(adj) <- list(V(gs)$label, V(gs)$label)
 
adj_long <- as.data.frame(as.table(adj)) %>%
  rename(author_x = Var1, author_y = Var2, weight = Freq) %>%
  mutate(
    author_x = factor(as.character(author_x), levels = lev),
    author_y = factor(as.character(author_y), levels = rev(lev)),
    # NA leaves non-collaborations blank rather than shading them as "zero"
    weight   = ifelse(weight == 0, NA_real_, weight)
  )

# --- Community boundaries as separator lines -------------------------------
boundaries <- cumsum(rle(comm_order)$lengths)
boundaries <- boundaries[-length(boundaries)]
 
# --- Plot -------------------------------------------------------------------
matrix_plot <- ggplot(adj_long, aes(x = author_x, y = author_y, fill = weight)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_vline(xintercept = boundaries + 0.5, colour = "grey25", linewidth = 0.6) +
  geom_hline(yintercept = length(lev) - boundaries + 0.5,
             colour = "grey25", linewidth = 0.6) +
  scale_fill_gradient(
    low = "#DCE9F5", high = "#1F4E79",
    na.value = "grey96", name = "Collaboration\nweight"
  ) +
  coord_fixed() +
  labs(
    title    = "Adjacency Matrix of the Top 25 Contributors",
    subtitle = "Rows and columns ordered by Louvain community, then degree;\ndark lines mark community boundaries",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(size = 10),
    axis.text.x     = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    axis.text.y     = element_text(size = 8),
    panel.grid      = element_blank(),
    legend.position = "right"
  )
 
print(matrix_plot)
 
ggsave("fig_adjacency_matrix.png", matrix_plot,
       width = 9, height = 8, dpi = 300, bg = "white")

```

# 9. Adjacency matrix, alphabetically ordered (for comparison)
```{r}
# Renders the SAME data unordered. Showing both proves that structure in a
# matrix view comes from the seriation, not the data - a strong point for the
# "comparison with other methods" section of the report.
 
adj_unordered <- adj_long %>%
  mutate(
    author_x = factor(as.character(author_x), levels = sort(lev)),
    author_y = factor(as.character(author_y), levels = rev(sort(lev)))
  )
 
matrix_unordered_plot <- ggplot(adj_unordered,
                                aes(x = author_x, y = author_y, fill = weight)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  scale_fill_gradient(low = "#DCE9F5", high = "#1F4E79",
                      na.value = "grey96", name = "Collaboration\nweight") +
  coord_fixed() +
  labs(
    title    = "The Same Matrix, Ordered Alphabetically",
    subtitle = "Without community-based ordering the block structure disappears",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 10),
    axis.text.x   = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    axis.text.y   = element_text(size = 8),
    panel.grid    = element_blank()
  )
 
print(matrix_unordered_plot)
 
ggsave("fig_adjacency_unordered.png", matrix_unordered_plot,
       width = 9, height = 8, dpi = 300, bg = "white")

```

# Adjacency matrix for Comparison side-by-side
```{r}
# install.packages("patchwork")
library(patchwork)

combined_matrix <- (
  matrix_plot        + labs(title = "(a) Ordered by Louvain community", subtitle = NULL) +
  matrix_unordered_plot + labs(title = "(b) Ordered alphabetically",   subtitle = NULL)
) &
  theme(plot.title  = element_text(face = "bold", size = 13),
        axis.text   = element_text(size = 6),
        legend.position = "right")

combined_matrix <- combined_matrix + plot_layout(guides = "collect")

ggsave("fig_matrix_comparison.png", combined_matrix,
       width = 15, height = 7.5, dpi = 300, bg = "white")

```

